---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gh

<!-- badges: start -->
[![R-CMD-check](https://github.com/r-lib/gh/workflows/R-CMD-check/badge.svg)](https://github.com/r-lib/gh/actions)
[![Codecov test coverage](https://codecov.io/gh/r-lib/gh/branch/master/graph/badge.svg)](https://codecov.io/gh/r-lib/gh?branch=master)
[![](http://www.r-pkg.org/badges/version/gh)](http://www.r-pkg.org/pkg/gh)
[![CRAN RStudio mirror downloads](http://cranlogs.r-pkg.org/badges/gh)](http://www.r-pkg.org/pkg/gh)
<!-- badges: end -->

Minimalistic client to access GitHub's
[REST](https://docs.github.com/rest) and [GraphQL](https://docs.github.com/graphql) APIs.

## Installation

Install the package from CRAN as usual:

```{r eval = FALSE}
install.packages("gh")
```

## Usage

```{r}
library(gh)
```

Use the `gh()` function to access all API endpoints. The endpoints are
listed in the [documentation](https://docs.github.com/rest).

The first argument of `gh()` is the endpoint. You can just copy and paste the 
API endpoints from the documentation. Note that the leading slash
must be included as well. 

From <https://docs.github.com/rest/reference/repos#list-repositories-for-a-user> you can copy and paste `GET /users/{username}/repos` into your `gh()`
call. E.g.

```{r}
my_repos <- gh("GET /users/{username}/repos", username = "gaborcsardi")
vapply(my_repos, "[[", "", "name")
```

The JSON result sent by the API is converted to an R object.

Parameters can be passed as extra arguments. E.g.

```{r}
my_repos <- gh(
  "/users/{username}/repos",
  username = "gaborcsardi",
  sort = "created")
vapply(my_repos, "[[", "", "name")
```

### POST, PATCH, PUT and DELETE requests

POST, PATCH, PUT, and DELETE requests can be sent by including the
HTTP verb before the endpoint, in the first argument. E.g. to
create a repository:

```{r eval = FALSE}
new_repo <- gh("POST /user/repos", name = "my-new-repo-for-gh-testing")
```

and then delete it:

```{r eval = FALSE}
gh("DELETE /repos/{owner}/{repo}", owner = "gaborcsardi",
   repo = "my-new-repo-for-gh-testing")
```

### Tokens

By default the `GITHUB_PAT` environment variable is used. Alternatively, 
one can set the `.token` argument of `gh()`.

### Pagination

Supply the `page` parameter to get subsequent pages:

```{r}
my_repos2 <- gh("GET /orgs/{org}/repos", org = "r-lib", page = 2)
vapply(my_repos2, "[[", "", "name")
```

## Environment Variables

* The `GITHUB_API_URL` environment variable is used for the default github
  api url.
* One of `GITHUB_PAT` or `GITHUB_TOKEN` environment variables is used, in
  this order, as default token.


## License

MIT © Gábor Csárdi, Jennifer Bryan, Hadley Wickham
