
```{r, setup, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  tidy = FALSE,
  error = FALSE,
  fig.width = 8,
  fig.height = 8)
```

# gh

> GitHub API

[![Linux Build Status](https://travis-ci.org/r-pkgs/gh.svg?branch=master)](https://travis-ci.org/r-pkgs/gh)
[![Windows Build status](https://ci.appveyor.com/api/projects/status/github/r-pkgs/gh?svg=true)](https://ci.appveyor.com/project/gaborcsardi/gh)
[![](http://www.r-pkg.org/badges/version/gh)](http://www.r-pkg.org/pkg/gh)
[![CRAN RStudio mirror downloads](http://cranlogs.r-pkg.org/badges/gh)](http://www.r-pkg.org/pkg/gh)


Minimalistic client to access
[GitHub's API v3](https://developer.github.com/v3/).

## Installation

```{r eval = FALSE}
devtools::install_github("r-pkgs/gh")
```

## Usage

```{r}
library(gh)
```

Use the `gh()` function to access all API endpoints. The endpoints are
listed in the [documentation](https://developer.github.com/v3/).

The first argument of `gh()` is the endpoint. Note that the leading slash
must be included as well. Parameters can be passed as extra arguments. E.g.

```{r}
my_repos <- gh("/user/repos", type = "public")
vapply(my_repos, "[[", "", "name")
```

The JSON result sent by the API is converted to an R object.

If the end point itself has parameters, these can also be passed
as extra arguments:

```{r}
j_repos <- gh("/users/:username/repos", username = "jeroenooms")
vapply(j_repos, "[[", "", "name")
```

### POST, PATCH, PUT and DELETE requests

POST, PATCH, PUT, and DELETE requests can be sent by including the
HTTP verb before the endpoint, in the first argument. E.g. to
create a repository:

```{r eval = FALSE}
new_repo <- gh("POST /user/repos", name = "my-new-repo-for-gh-testing")
```

and then delete it:

```{r eval = FALSE}
gh("DELETE /repos/:owner/:repo", owner = "gaborcsardi",
   repo = "my-new-repo-for-gh-testing")
```

### Tokens

By default the `GITHUB_PAT` environment variable is used. Alternatively, 
one can set the `.token` argument of `gh()`.

### Pagination

Supply the `page` parameter to get subsequent pages:

```{r}
my_repos2 <- gh("GET /users/:username/repos", username = "gaborcsardi",
  type = "public", page = 2)
vapply(my_repos2, "[[", "", "name")
```

## License

MIT © [Gábor Csárdi](https://github.com/gaborcsardi).
