---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gh

<!-- badges: start -->
[![R-CMD-check](https://github.com/r-lib/gh/workflows/R-CMD-check/badge.svg)](https://github.com/r-lib/gh/actions)
[![](https://www.r-pkg.org/badges/version/gh)](https://www.r-pkg.org/pkg/gh)
[![CRAN Posit mirror downloads](https://cranlogs.r-pkg.org/badges/gh)](https://www.r-pkg.org/pkg/gh)
[![R-CMD-check](https://github.com/r-lib/gh/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/r-lib/gh/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/r-lib/gh/graph/badge.svg)](https://app.codecov.io/gh/r-lib/gh)
<!-- badges: end -->

Minimalistic client to access GitHub's
[REST](https://docs.github.com/rest) and [GraphQL](https://docs.github.com/graphql) APIs.

## Installation and setup

Install the package from CRAN as usual:

```{r}
#| eval: false
install.packages("gh")
```

Install the development version from GitHub:

```{r}
#| eval: false
pak::pak("r-lib/gh")
```

### Authentication

The value returned by `gh::gh_token()` is used as Personal Access Token
(PAT). A token is needed for some requests, and to help with rate limiting.
gh can use your regular git credentials in the git credential store, via
the gitcreds package. Use `gitcreds::gitcreds_set()` to put a PAT into the
git credential store. If you cannot use the credential store, set the
`GITHUB_PAT` environment variable to your PAT. See the details in the
`?gh::gh_token` manual page and the manual of the gitcreds package.

### API URL

* The `GITHUB_API_URL` environment variable, if set, is used for the default github api url.

## Usage

```{r}
library(gh)
```

Use the `gh()` function to access all API endpoints. The endpoints are
listed in the [documentation](https://docs.github.com/rest).

The first argument of `gh()` is the endpoint. You can just copy and paste the
API endpoints from the documentation. Note that the leading slash
must be included as well.

From <https://docs.github.com/rest/reference/repos#list-repositories-for-a-user> you can copy and paste `GET /users/{username}/repos` into your `gh()`
call. E.g.

```{r}
my_repos <- gh("GET /users/{username}/repos", username = "gaborcsardi")
vapply(my_repos, "[[", "", "name")
```

The JSON result sent by the API is converted to an R object.

Parameters can be passed as extra arguments. E.g.

```{r}
my_repos <- gh(
  "/users/{username}/repos",
  username = "gaborcsardi",
  sort = "created")
vapply(my_repos, "[[", "", "name")
```

### POST, PATCH, PUT and DELETE requests

POST, PATCH, PUT, and DELETE requests can be sent by including the
HTTP verb before the endpoint, in the first argument. E.g. to
create a repository:

```{r}
#| eval: false
new_repo <- gh("POST /user/repos", name = "my-new-repo-for-gh-testing")
```

and then delete it:

```{r}
#| eval: false
gh("DELETE /repos/{owner}/{repo}", owner = "gaborcsardi",
   repo = "my-new-repo-for-gh-testing")
```

### Tokens

By default the `GITHUB_PAT` environment variable is used. Alternatively,
one can set the `.token` argument of `gh()`.

### Pagination

Supply the `page` parameter to get subsequent pages:

```{r}
my_repos2 <- gh("GET /orgs/{org}/repos", org = "r-lib", page = 2)
vapply(my_repos2, "[[", "", "name")
```

## Environment Variables

* The `GITHUB_API_URL` environment variable is used for the default github
  api url.
* The `GITHUB_PAT` and `GITHUB_TOKEN` environment variables are used, if
  set, in this order, as default token. Consider using the git credential
  store instead, see `?gh::gh_token`.

## Code of Conduct

Please note that the gh project is released with a
[Contributor Code of Conduct](https://gh.r-lib.org/CODE_OF_CONDUCT.html).
By contributing to this project, you agree to abide by its terms.

## License

MIT © Gábor Csárdi, Jennifer Bryan, Hadley Wickham
