---
title: "Managing Personal Access Tokens"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Managing Personal Access Tokens}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

*NOTE: This aspect of gh is under active development, so there may be brief periods where things are out of sync.
Please bear with and feel free to let us know in a GitHub issue.*

```{r setup}
library(gh)
```

<!-- This vignette uses a convention of "one sentence per line" in prose. -->

gh generally sends a Personal Access Token (PAT) with its requests.
Some endpoints of the GitHub API can be accessed without authenticating yourself.
But once your API use becomes more frequent, you will want a PAT to prevent problems with rate limits and to access all possible endpoints.

This article describes how to store your PAT, so that gh can find it (automatically, in most cases). The function gh uses for this is `gh_token()`.

More resources on how and why to get a PAT:

  * GitHub documentation on [Creating a personal access token](https://docs.github.com/en/github/authenticating-to-github/creating-a-personal-access-token)
  * In the usethis package:
    - [Setup advice re: PATs](https://usethis.r-lib.org/articles/articles/usethis-setup.html#get-and-store-a-github-personal-access-token-1)
    - `usethis::create_github_token()` guides you through the process of getting
      a new PAT
  * `credentials::set_github_pat()`
  
## PAT and host

`gh::gh()` allows the user to provide a PAT via the `.token` argument and to specify a host other than "github.com" via the `.api_url` argument.
(Some companies and universities run their own instance of GitHub Enterprise.)

```{r, eval = FALSE}
gh(endpoint, ..., .token = NULL, ..., .api_url = NULL, ...)
```

However, it's annoying to always provide your PAT or host and it's unsafe for your PAT to appear explicitly in your R code.
It's important to make it *possible* for the user to provide the PAT and/or API URL directly, but it should rarely be necessary.
`gh::gh()` is designed to play well with more secure, less fiddly methods for expressing what you want.

How are `.api_url` and `.token` determined when the user does not provide them?

  1. `.api_url` defaults to the value of the `GITHUB_API_URL` environment
    variable and, if that is unset, falls back to `"https://api.github.com"`.
    This is always done before worrying about the PAT.
  1. The PAT is obtained via a call to `gh_token(.api_url)`. That is, the token
    is looked up based on the host.
    
`gh_token()` actually has a second argument, `strategy`:

```{r, eval = FALSE}
gh_token(api_url = NULL, strategy = NULL)
```

The PAT corresponding to `api_url` is searched for with a `strategy` that
looks in one or more of these places, usually in this order:

* "env": environment variable(s)
* "git": Git credential store
* "key": OS-level keychain (requires the keyring package)

## PAT in an environment variable

The "env" search strategy looks for a PAT in specific environment variables.
If `api_url` targets "github.com", these variables are consulted, in order:

1. `GITHUB_PAT_GITHUB_COM`
1. `GTIHUB_PAT`
1. `GITHUB_TOKEN`

If `api_url` targets another GitHub deployment, such as "github.acme.com", this variable is consulted:

* `GITHUB_PAT_GITHUB_ACME_COM`

In both cases, the suffix in `GITHUB_PAT_<SUFFIX>` is derived from `api_url` using the helper `slugify_url()`.

Looking up the PAT in an environment variable is definitely more secure than including it explicitly in your code, i.e. providing via `gh(token = "xyz")`.
The simplest way to set this up is to define, e.g., `GITHUB_PAT` in your `.Renviron` startup file.
This is the entry-level solution.

However, ideally you would not store your PAT in plain text like this.
It is also undesirable to make your PAT available to all your R sessions, regardless of actual need.
Both make it more likely you will expose your PAT publicly, by accident.
Therefore, it is strongly recommended to store your PAT in the Git credential store or system keychain and allow gh to retrieve it on-demand.
See the next two sections for more.

## PAT in the Git credential store

The "git" search `strategy` looks up the PAT corresponding to `api_url` in the Git credential store.
This `strategy` has the advantage of using official Git tooling, specific to your operating system, for managing secrets.
The first time the "git" `strategy` is invoked, you may be prompted for your PAT and, if it validates, it is stored for future re-use with this `api_url`.
For the remainder of the current R session, the PAT is also available via one of the usual environment variables:

* `GITHUB_PAT` for "github.com"
* `GITHUB_PAT_GITHUB_ACME_COM` for "github.acme.com"

This pattern of retrieving the PAT from the store upon first need and caching it in an environment variable is why the default `strategy` tries "env" and then "git":

1. The initial "env" search fails.
2. The "git" search succeeds and sets an environment variable in the session.
3. Subsequent "env" searches succeed.

*TO BE FILLED IN: How to get your PAT into the Git credential store.*

## PAT in the system keyring:

The "key" search `strategy` uses the Suggested keyring package to retrieve your PAT from the system keyring, on Windows, macOS and Linux.
To activate keyring, specify a `strategy` that includes "key" or set the `GH_KEYRING` environment variable to `true`, e.g. in your `.Renviron` file.
The keys queried for a PAT are exactly the same as the environment variable names consulted for the "env" `strategy`. For "github.com", the first keyring check looks like this:

```r
keyring::key_get("GITHUB_PAT_GITHUB_COM")
```

gh uses the default keyring backend and the default keyring within that backend.
See `keyring::default_backend()` for details and changing these defaults.

If the selected keyring is locked, and the session is interactive, then gh will try to unlock it.
If the keyring is locked, and the session is not interactive, then gh will not use the keyring.
Note that some keyring backends cannot be locked, e.g. the one that uses environment variables.
On some OSes, e.g. typically on macOS, you need to allow R to access the system keyring.
You can allow this separately for each access, or for all future accesses, until you update or re-install R.
You typically need to give access to each R GUI (e.g. RStudio) and the command line R program separately.
To store your PAT on the keyring, run:

```r
keyring::key_set("GITHUB_PAT")
```

## Recommendations

For a normal user, on a machine used for interactive development, we recommend:

  * Store your PAT(s) in an official credential store or keychain for your
    OS.
  * Do **not** store your PAT(s) in plain text in, e.g., `.Renviron`. In the
    past, this has been a common and recommended practice for pragmatic reasons.
    However, credentials/keyring/gh have now evolved to the point where it's
    possible for all of us to follow better security practices.
  * If you use a general-purpose password manager, like 1Password or LastPass,
    you may *also* want to store your PAT(s) there. Why? If your PAT is
    "forgotten" from the OS-level credential store, intentionally or not, you'll
    need to provide it again when prompted. For security reasons, none of the
    tools involved will help you re-discover a PAT.
    
    If you don't have any other record of your PAT, you'll have to get a new
    PAT whenever this happens. This is not the end of the world. But if you
    aren't disciplined about deleting lost PATs from
    <https://github.com/settings/tokens>, you will eventually find yourself in a
    confusing situation where you can't be sure which PAT(s) are in use.

On a headless system, such as on a CI/CD platform, provide the necessary PAT(s) via secure environment variables. Regular environment variables can be used to configure less sensitive settings, such as the API host.
Don't expose your PAT by doing something silly like dumping all environment variables to a log file.

Note that on GitHub Actions, specifically, a personal access token is [automatically available to the workflow](https://docs.github.com/en/actions/configuring-and-managing-workflows/authenticating-with-the-github_token) as the `GITHUB_TOKEN` secret.
That is why many workflows in the R community contain this snippet:

``` yaml
env:
  GITHUB_PAT: ${{ secrets.GITHUB_TOKEN }}
```

This makes the automatic PAT available as the `GITHUB_PAT` environment variable.
If that PAT doesn't have the right permissions, then you'll need to explicitly provide one that does (see link above for more).

## Failure

If there is no PAT to be had, `gh::gh()` sends a request with no token.
(Internally, the `Authorization` header is omitted if the PAT is found to be the empty string, `""`.)

What do PAT-related failures look like?

If no PAT is sent and the endpoint requires no auth, the request probably succeeds!
At least until you run up against rate limits.
If the endpoint requires auth, you'll get an HTTP error, possibly this one:

```
GitHub API error (401): 401 Unauthorized
Message: Requires authentication
```

If a PAT is first discovered in an environment variable, it is taken at face value.
The two most common ways to arrive here are PAT specification via `.Renviron` or as a secret in a CI/CD platform, such as GitHub Actions.
If the PAT is invalid, the first affected request will fail, probably like so:

```
GitHub API error (401): 401 Unauthorized
Message: Bad credentials
```

This will also be the experience if an invalid PAT is provided directly via `.token`.

If the PAT is retrieved from the Git credential store or is elicited from the user via prompt, the PAT is immediately and explicitly checked for basic validity.

Even a valid PAT can lead to a downstream error, if it has insufficient scopes with respect to a specific request.

## API URL specifications

Above we explained that environment variables are consulted during token lookup and that their names are based on the target host.
The table below shows which environment variables are consulted when targeting "github.com" or "github.uni.edu", a fictional instance of GitHub Enterprise hosted by a university.

Environment variables that depart from the `<LABEL>_<SLUG>` pattern are included for compatibility with credentials (\*) or to uphold a convention that's been adopted by the wider package ecosystem (\*\*).

*TODO: Maybe credentials is new enough that we can eliminate `GITHUB_PAT_USER` in favor of `GITHUB_USER_GITHUB_COM`? Even making `GITHUB_USER` an acceptable short form would be more consistent with the overall pattern.*

| URL from user or<br>GITHUB_API_URL                            | host<br>SLUG = slugify_host(host) | Env var for user:<br>GITHUB_USER_\<SLUG\>     | Env var for PAT:<br>GITHUB_PAT_\<SLUG\>                       |
|---------------------------------------------------------------|-----------------------------------|-----------------------------------------------|---------------------------------------------------------------|
| https://api.github.com<br>OR<br>https://github.com            | github.com<br>GITHUB_COM          | GITHUB_USER_GITHUB_COM<br>GITHUB_PAT_USER (\*) | GITHUB_PAT_GITHUB_COM<br>GITHUB_PAT (\*\*)<br>GITHUB_TOKEN (\*\*) |
| https://github.uni.edu/api/v3<br>OR<br>https://github.uni.edu | github.uni.edu<br>GITHUB_UNI_EDU  | GITHUB_USER_GITHUB_UNI_EDU                    | GITHUB_PAT_GITHUB_UNI_EDU                                     |

What is this column about the *user*? That hasn't been explained yet.
This is related to how credentials are retrieved from the Git credential store.
The TL;DR is that most users should not worry about this and accept default behaviour.

### Role of `username` for the Git credential store

When one interacts with the Git credential store, accessed via [`git credential <fill|approve|reject>`](https://git-scm.com/docs/git-credential), it's necessary to describe the credential of interest.

Suppressing a lot of detail, a key for each credential is built from:

  * `host`: The remote hostname for a network credential.
  * `username`: The credential’s username.

When using a PAT, GitHub doesn't actually consult or need the `username`.
It's implicit in the PAT.
And yet `username` is still a necessary part of a credential's key.
The credentials package follows a convention from [Git Credential Manager Core](https://github.blog/2020-07-02-git-credential-manager-core-building-a-universal-authentication-experience/) and, by default, uses `"PersonalAccessToken"` as the `username` for all PATs.
There is a good reason to *not* use the user's actual `username`. If the user has already stored a credential through normal HTTPS Git operations, we don't want to clobber it with the PAT.

Neither credentials nor gh makes it easy to explicitly store a PAT with a particular `username`.
The environment variable `GITHUB_USER_<SLUG>` already provides some access, if it is defined prior to a PAT prompt.
The `username` could provide a method for working with more than one PAT for a given host, perhaps for different users and/or with different scopes.
If needed, such workflows could be built out in the future.
